---
title: "week_9_lab_tasks"
output: html_document
---

table id: 123518
stats NZ table url: https://datafinder.stats.govt.nz/table/123518-geographic-areas-table-2026/

Statistical Area 2 2026,
layer id: 123515


https://koordinates.com/services/query/v1/vector.json?key=c301ae7a8cea42cfa1a558f4838a53a5&layer=3936&x=172.6398&y=43.5256&max_results=3&radius=20000&geometry=true&with_field_names=true



Some old code to delete later
url = "https://koordinates.com/services/query/v1/vector.json?key=c301ae7a8cea42cfa1a558f4838a53a5&layer=123515&x=172.6398&y=-43.5256&max_results=3&radius=20000&geometry=true&with_field_names=true"

response = requests.get(url)

#Check if the request was successful (Status Code 200)
if response.status_code == 200:
  data = response.json()
  print("Success!")
  print(data)
else:
    print(f"Failed to fetch data. Status code: {response.status_code}")

#to install stuff
reticulate::py_require(c("requests", "pandas"))


## Querying the API
Queries the Koordinates API. We give it a longitude (x) and 
latitude (y) and it sends back whatever data matches that location
If the request fails for some reason (bad internet, server 
down, etc) it just returns None

```{python}
import os
import time
import requests
import pandas as pd
from concurrent.futures import ThreadPoolExecutor, as_completed

# --- CONFIGURATION CONTROLS ---
API_KEY = "c301ae7a8cea42cfa1a558f4838a53a5"
LAYER_ID = 123515
MAX_WORKERS = 8  # Number of parallel workers querying the API simultaneously
RADIUS = 20000

def query_layer(x, y, max_results, layer_id=LAYER_ID, radius=RADIUS, api_key=API_KEY):
    url = (
        f"https://koordinates.com/services/query/v1/vector.json"
        f"?key={api_key}&layer={layer_id}&x={x}&y={y}"
        f"&max_results={max_results}&radius={radius}&geometry=true&with_field_names=true"
    )
    response = requests.get(url)
    if response.status_code == 200:
        print("y")
        return response.json()
    else:
        print(f"Failed to fetch data for ({x}, {y}). Status code: {response.status_code}")
        print("g")
        return None

# checking the query layer works
query_layer(172.6398,-43.5256,  3)

```

Column we are getting: SA22026_V1_00. <-- may change this later????

## Getting the info (area code + suburb name) for the location
Uses the query function above but pulls out just the area code and the name. 
```{python}
def get_area_info(x, y):
    """returns SA22026_V1_00 value (area code) and SA22026_V1_00_NAME (suburb name) for latitude and      longitude, or returns None, None"""
    # Query the API
    data = query_layer(x, y, max_results=1)

    # query_layer() returns None if the request failed
    if data is None:
        return None, None

    try:
        features = data["vectorQuery"]["layers"][str(LAYER_ID)]["features"]

        # If no matches found at this point return None
        if len(features) == 0:
            return None, None

        # Get the area code and suburb name from the result's properties
        props = features[0]["properties"]
        return props["SA22026_V1_00"], props["SA22026_V1_00_NAME"]

    # do not run for key error (where field name doesn't exist) or index error (features list was         empty and not caught by len()==0 bit)
    except (KeyError, IndexError):
        return None, None
```

A couple tests
```{python}
code, name = get_area_info(172.5810, -43.5227)
print(name)

code, name = get_area_info(172.6362, -43.5320)  # Cathedral Square, Christchurch
print(code, name)
```
```{python}
# Define input/output file paths, directories, and configuration IDs for week 9 datasets.
from pathlib import Path
import numpy as np

DATA_DIR = Path("week_9_datasets")
AIRBNB_IN = DATA_DIR / "airbnb_cleaned_week_8.csv"
TENANCY_IN = DATA_DIR / "tenancy_cleaned_week_8.csv"
AIRBNB_WITH_AREA_OUT = DATA_DIR / "airbnb_with_area_codes.csv"
GEO_AREAS_TABLE_CSV = DATA_DIR / "geographic_areas_table_2026.csv"
GEO_AREAS_TABLE_ID = 123518
```

```{python}
# Perform parallel batch geocoding for unique coordinates using a cache to skip existing locations, then save the enriched dataset.

def run_batch_geocoding(airbnb: pd.DataFrame) -> pd.DataFrame:
    unique_coords = airbnb[["latitude", "longitude"]].drop_duplicates().reset_index(drop=True)

    if AIRBNB_WITH_AREA_OUT.exists(): # If airbnb_with_area_codes.csv exist, every future call to this function loads it first and only queries coordinates that are still missing an area code. If everything geocoded successfully, to do will be empty.
        cached = pd.read_csv(AIRBNB_WITH_AREA_OUT)
        done = cached.dropna(subset=["area_code"])[["latitude", "longitude", "area_code", "area_name"]].drop_duplicates()
        merged = unique_coords.merge(done, on=["latitude", "longitude"], how="left")
    else:
        merged = unique_coords.copy()
        merged["area_code"] = np.nan
        merged["area_name"] = np.nan

    todo = merged[merged["area_code"].isna()]
    print(f"{len(todo)} / {len(unique_coords)} unique coordinates still need querying")

    if len(todo) > 0:
        records = []
        with ThreadPoolExecutor(max_workers=MAX_WORKERS) as executor:
            futures = {
                executor.submit(get_area_info, row.longitude, row.latitude): (row.latitude, row.longitude) # calls Alisha's get_area_info()
                for row in todo.itertuples()
            }
            completed = 0
            for future in as_completed(futures):
                lat, lon = futures[future]
                area_code, area_name = future.result()
                records.append({"latitude": lat, "longitude": lon, "area_code": area_code, "area_name": area_name})
                completed += 1
                if completed % 200 == 0:
                    print(f"{completed} / {len(todo)} done")

        results_df = pd.DataFrame(records)
        merged = merged.merge(results_df, on=["latitude", "longitude"], how="left", suffixes=("", "_new"))
        merged["area_code"] = merged["area_code"].combine_first(merged["area_code_new"])
        merged["area_name"] = merged["area_name"].combine_first(merged["area_name_new"])
        merged = merged.drop(columns=["area_code_new", "area_name_new"])

    airbnb_out = airbnb.merge(merged, on=["latitude", "longitude"], how="left")
    airbnb_out.to_csv(AIRBNB_WITH_AREA_OUT, index=False)
    print(f"Saved {AIRBNB_WITH_AREA_OUT} ({airbnb_out['area_code'].notna().sum()} / {len(airbnb_out)} rows matched)")
    return airbnb_out
```

```{python}
# Load the cleaned Airbnb dataset from CSV and run batch geocoding to assign geographic area information.
airbnb = pd.read_csv(AIRBNB_IN)
airbnb_geocoded = run_batch_geocoding(airbnb)
```
```{r}
# Look at the now airbnb file that has area codes in it got from Koordinates
data <- read_csv('week_9_datasets/airbnb_with_area_codes.csv')
summary(data)
```
```{r}
# double check has area code in it 
glimpse(data)
```
```{r}
# Merge the airbnb_with_area_codes with tenancy_cleaned_week_8 using left join to keep all the columns. 
library(tidyverse)
library(lubridate)

airbnb <- read_csv("week_9_datasets/airbnb_with_area_codes.csv")
tenancy <- read_csv("week_9_datasets/tenancy_cleaned_week_8.csv")

airbnb <- airbnb %>%
  mutate(
    month_date = ym(month_year),
    quarter = floor_date(month_date, unit = "quarter"),
    area_code = as.numeric(area_code)
  )

tenancy_all <- tenancy %>%
  filter(
    `Dwelling Type` == "ALL",
    `Number Of Beds` == "ALL",
    `Location Id` != -99
  ) %>%
  rename(area_code = `Location Id`) %>%
  mutate(quarter = as.Date(TimeFrame))

# Left join
joined <- airbnb %>%
  left_join(tenancy_all, by = c("area_code", "quarter"))

# save to a new dataset: airbnb_tenancy_joined.csv
write_csv(joined, "week_9_datasets/airbnb_tenancy_joined.csv")
```
```{r}
joined_data <- read_csv('week_9_datasets/airbnb_tenancy_joined.csv')
glimpse(joined_data)
```
17/9/2026 - Alice
•	Dropped rows with location ID -99 from the Week 8 tenancy file.

•	Created week_9_datasets folder, which contains the two cleaned Week 8 files: Airbnb_cleaned_week_8 and tenancy_cleaned_week_8.

•	Airbnb data has latitude/longitude, while tenancy data has area codes. Used the Koordinates API to geocode each unique Airbnb coordinate to its official SA2 area code.

•	Downloaded geographic-areas-tables-2026.csv (originally a zip file with 5 files in it, including geographic-areas-table-2026.csv) from Koordinates

•	Created run_batch_geocoding(), which uses Alisha’s get_area_info() and query_layer() functions. It batch-geocoded ~4,000 unique coordinates. The function saves the result as airbnb_with_area_code.csv with a new area_code column. It also checks for existing output and only queries coordinates with missing area codes, avoiding unnecessary API calls from now on.

•	Joined the geocoded Airbnb data with tenancy_cleaned_week_8 using a left join, keeping all Airbnb listings. It first converted Airbnb’s monthly dates into quarterly periods to match the rental data, then joined on both area_code and quarter.
Used a left join so Airbnb listings are retained even when there is no matching rental data for their area/quarter. (Inner join will delete any data without matching rental data. Full/outer join will include tenancy rows that are NOT in chch, which is not meaningful)

•	Saved the final combined dataset as airbnb_tenancy_joined.csv.

Note merged data assigned variable name: joined_data 
Heads up airbnb_tenancy_joined.csv is the finished joined data: joined_data <- read_csv('week_9_datasets/airbnb_tenancy_joined.csv'. So mess with joined_data however you want for your part. Just not write_csv() back to that exact filename (hopefully we won't have to write another csv file but just in case) otherwise will overwrite it for everyone. If we need to save our own version, give it a totally different name, like price_gap.csv or something so it doesn't get confused with the original.

# What’s the median AirBnB price in Christchurch Central (Location ID 326600) - Alice

```{r}
task1_data <- joined_data
```

# In which part of Christchurch can we observe the craziest (largest) gap between short- and long-term rental prices? - Charlie or Alisha 

# Compare how many AirBnBs and rental properties we have in each location - Charlie or Alisha
