---
title: "week_9_lab_tasks"
output: html_document
---

table id: 123518
stats NZ table url: https://datafinder.stats.govt.nz/table/123518-geographic-areas-table-2026/
Statistical Area 2 2026,
layer id: 123515

https://koordinates.com/services/query/v1/vector.json?key=c301ae7a8cea42cfa1a558f4838a53a5&layer=3936&x=172.6398&y=43.5256&max_results=3&radius=20000&geometry=true&with_field_names=true

#Setting up the environment installing packages
```{r}
# This line sets up the Python environment automatically - must run first, don't delete
# packages live in a local cache on each machine, not in this file, so this line tells reticulate which packages to use/build each fresh session.
reticulate::py_require(c("pandas", "requests", "numpy"))
```

## Querying the API
Queries the Koordinates API. 
We give it a longitude (x) and latitude (y) and it sends back data that matches that location. If the request fails for some reason (bad internet, server down, etc) it just returns None
```{python}
# importing packages
import os
import time
import requests
import pandas as pd
from concurrent.futures import ThreadPoolExecutor, as_completed

# --- CONFIGURATION CONTROLS ---
API_KEY = "c301ae7a8cea42cfa1a558f4838a53a5"
LAYER_ID = 123515
MAX_WORKERS = 8  # Number of parallel workers querying the API simultaneously
RADIUS = 20000 # search radius in metres around the x,y point when querying the API

def query_layer(x, y, max_results, layer_id=LAYER_ID, radius=RADIUS, api_key=API_KEY):
  """Query the Koordinates API for features near a given coordinate 
  and return the JSON response, or None if the request failed."""

    url = (
        f"https://koordinates.com/services/query/v1/vector.json"
        f"?key={api_key}&layer={layer_id}&x={x}&y={y}"
        f"&max_results={max_results}&radius={radius}&geometry=true&with_field_names=true"
    )
    response = requests.get(url)
    if response.status_code == 200: # if the request is sucessful
        print("y")
        return response.json()
    # if unsucessful return None
    else:
        print(f"Failed to fetch data for ({x}, {y}). Status code: {response.status_code}")
        print("g")
        return None

# checking the query layer works
query_layer(172.6398,-43.5256,  3)

```

## Getting the info (area code + suburb name) for the location
Uses the query function above but pulls out just the area code and the name. 
```{python}
def get_area_info(x, y):
    """returns SA22026_V1_00 value (area code) and SA22026_V1_00_NAME (suburb name) for latitude and      longitude, or returns None, None"""
    # Query the API
    data = query_layer(x, y, max_results=1)

    # query_layer() returns None if the request failed
    if data is None:
        return None, None

    try:
        features = data["vectorQuery"]["layers"][str(LAYER_ID)]["features"]

        # If no matches found at this point return None
        if len(features) == 0:
            return None, None

        # Get the area code and suburb name from the result's properties
        props = features[0]["properties"]
        return props["SA22026_V1_00"], props["SA22026_V1_00_NAME"]

    # do not run for key error (where field name doesn't exist) or index error (features list was         empty and not caught by len()==0 bit)
    except (KeyError, IndexError):
        return None, None
```

A couple tests
```{python}
# lookig at area names
code, name = get_area_info(172.5810, -43.5227)
print(name)

code, name = get_area_info(172.6362, -43.5320)  # Cathedral Square, Christchurch
print(code, name)
```
Defining things
```{python}
# Define input/output file paths, directories, and configuration IDs for week 9 datasets.
from pathlib import Path
import numpy as np

DATA_DIR = Path("week_9_datasets")
AIRBNB_IN = DATA_DIR / "airbnb_cleaned_week_8.csv"
TENANCY_IN = DATA_DIR / "tenancy_cleaned_week_8.csv"
AIRBNB_WITH_AREA_OUT = DATA_DIR / "airbnb_with_area_codes.csv"
GEO_AREAS_TABLE_CSV = DATA_DIR / "geographic_areas_table_2026.csv"
GEO_AREAS_TABLE_ID = 123518
```

This function geocodes all the AirBnB listings at once, excluding duplicates. It looks at every unique lat/long pair, checks if there's already a saved file from a previous run (skips anything already done), then queries the API for whatever's left. Once finished, it merges the area codes back onto the full AirBnB dataset and saves it to CSV.
```{python}
# Perform parallel batch geocoding for unique coordinates using a cache to skip existing locations, then save the enriched dataset.
def run_batch_geocoding(airbnb: pd.DataFrame) -> pd.DataFrame:
    unique_coords = airbnb[["latitude", "longitude"]].drop_duplicates().reset_index(drop=True) # get lat/long dropping duplicates

# If airbnb_with_area_codes.csv exist, every future call to this function loads it first and only queries coordinates that are still missing an area code. If everything geocoded successfully, to do will be empty.
    if AIRBNB_WITH_AREA_OUT.exists(): 
        cached = pd.read_csv(AIRBNB_WITH_AREA_OUT)
        done = cached.dropna(subset=["area_code"])[["latitude", "longitude", "area_code", "area_name"]].drop_duplicates()
        merged = unique_coords.merge(done, on=["latitude", "longitude"], how="left")
    else:
        merged = unique_coords.copy()
        merged["area_code"] = np.nan
        merged["area_name"] = np.nan

    todo = merged[merged["area_code"].isna()]
    print(f"{len(todo)} / {len(unique_coords)} unique coordinates still need querying")

    if len(todo) > 0:
        records = []
        with ThreadPoolExecutor(max_workers=MAX_WORKERS) as executor:
            futures = {
                executor.submit(get_area_info, row.longitude, row.latitude): (row.latitude, row.longitude) # calls Alisha's get_area_info()
                for row in todo.itertuples()
            }
            completed = 0
            for future in as_completed(futures):
                lat, lon = futures[future]
                area_code, area_name = future.result()
                records.append({"latitude": lat, "longitude": lon, "area_code": area_code, "area_name": area_name})
                completed += 1
                if completed % 200 == 0:
                    print(f"{completed} / {len(todo)} done")

# merge freshly-queried results back into merged, keeping cached values where they already existed
        results_df = pd.DataFrame(records)
        merged = merged.merge(results_df, on=["latitude", "longitude"], how="left", suffixes=("", "_new"))
        merged["area_code"] = merged["area_code"].combine_first(merged["area_code_new"])
        merged["area_name"] = merged["area_name"].combine_first(merged["area_name_new"])
        merged = merged.drop(columns=["area_code_new", "area_name_new"])

# merge area codes back onto the full (non-deduplicated) airbnb dataset and save
    airbnb_out = airbnb.merge(merged, on=["latitude", "longitude"], how="left")
    airbnb_out.to_csv(AIRBNB_WITH_AREA_OUT, index=False)
    print(f"Saved {AIRBNB_WITH_AREA_OUT} ({airbnb_out['area_code'].notna().sum()} / {len(airbnb_out)} rows matched)")
    return airbnb_out
```


Loads the cleaned Airbnb CSV and runs it through run_batch_geocoding(), which adds the area code and name to every listing based on its coordinates
```{python}
# Load the cleaned Airbnb dataset from CSV and run batch geocoding to assign geographic area information.
airbnb = pd.read_csv(AIRBNB_IN)
airbnb_geocoded = run_batch_geocoding(airbnb)
```

```{r}
# Look at the now airbnb file that has area codes in it got from Koordinates
library(tidyverse)
library(lubridate)

data <- read_csv('week_9_datasets/airbnb_with_area_codes.csv')
summary(data)
```

```{r}
# double check has area code in it 
glimpse(data)
```

Joins the geocoded Airbnb data with the tenancy bond data on area_code and quarter, so each Airbnb listing gets matched with the long-term rental stats for its area and time period. 
It first converts Airbnb's monthly dates into quarters and makes sure area_code is numeric (to match tenancy's format), then filters the tenancy data down to just the overall totals (Dwelling Type = "ALL", Number Of Beds = "ALL") and drops rows with an invalid location ID (-99). 
A left join is used so every Airbnb listing is kept, even if there's no matching tenancy data for that area/quarter. The result is saved as airbnb_tenancy_joined.csv.
```{r}
# Merge the airbnb_with_area_codes with tenancy_cleaned_week_8 using left join to keep all the columns. 

airbnb <- read_csv("week_9_datasets/airbnb_with_area_codes.csv")
tenancy <- read_csv("week_9_datasets/tenancy_cleaned_week_8.csv")

airbnb <- airbnb %>%
  mutate(
    month_date = ym(month_year), # convert to a date
    quarter = floor_date(month_date, unit = "quarter"), # round down to match tenancy's quarters
    area_code = as.numeric(area_code) # match tenancy's format for the join
  )

tenancy_all <- tenancy %>%
  filter(
    `Dwelling Type` == "ALL", # Keeps only rows where the dwelling type is set to "ALL"
    `Number Of Beds` == "ALL", # Keeps only rows where the number of beds is set to "ALL"
    `Location Id` != -99 # Removes rows where the location ID is -99
  ) %>%
  rename(area_code = `Location Id`) %>%
  mutate(quarter = as.Date(TimeFrame))

# Left join on area_code and quarter, keeps every Airbnb row even without a match
joined <- airbnb %>%
  left_join(tenancy_all, by = c("area_code", "quarter"))

# save to a new dataset: airbnb_tenancy_joined.csv
write_csv(joined, "week_9_datasets/airbnb_tenancy_joined.csv")
```

```{r}
# load the finished joined dataset and check its structure
joined_data <- read_csv('week_9_datasets/airbnb_tenancy_joined.csv')
glimpse(joined_data)
```
17/9/2026 - Alice
•	Dropped rows with location ID -99 from the Week 8 tenancy file and filtered to keep only rows where Dwelling Type == "ALL" and Number Of Beds == "ALL". (This makes sure there is only 1 row per suburb and quarter, which prevents our Airbnb listings from getting duplicated when we merge the files.)

•	Created week_9_datasets folder, which contains the two cleaned Week 8 files: Airbnb_cleaned_week_8 and tenancy_cleaned_week_8.

•	Airbnb data has latitude/longitude, while tenancy data has area codes. Used the Koordinates API to geocode each unique Airbnb coordinate to its official SA2 area code.

•	Downloaded geographic-areas-tables-2026.csv (originally a zip file with 5 files in it, including geographic-areas-table-2026.csv) from Koordinates

•	Created run_batch_geocoding(), which uses Alisha’s get_area_info() and query_layer() functions. It batch-geocoded ~4,000 unique coordinates. The function saves the result as airbnb_with_area_codes.csv with a new area_code column. It also checks for existing output and only queries coordinates with missing area codes, avoiding unnecessary API calls from now on.

•	Joined the geocoded Airbnb data with tenancy_cleaned_week_8 using a left join, keeping all Airbnb listings. It first converted Airbnb’s monthly dates into quarterly periods to match the rental data, then joined on both area_code and quarter.
Used a left join so Airbnb listings are retained even when there is no matching rental data for their area/quarter. (Inner join will delete any data without matching rental data. Full/outer join will include tenancy rows that are NOT in chch, which is not meaningful)

•	Saved the final combined dataset as airbnb_tenancy_joined.csv.

Note merged data assigned variable name: joined_data 
Heads up airbnb_tenancy_joined.csv is the finished joined data: joined_data <- read_csv('week_9_datasets/airbnb_tenancy_joined.csv'. So mess with joined_data however you want for your part. Just not write_csv() back to that exact filename (hopefully we won't have to write another csv file but just in case) otherwise will overwrite it for everyone. If we need to save our own version, give it a totally different name, like price_gap.csv or something so it doesn't get confused with the original.

# What’s the median AirBnB price in Christchurch Central (Location ID 326600) - Alice

```{r}
task1_data <- joined_data

# Quick check: see what area_code actually looks like (numbers? text? are there NAs?)
glimpse(task1_data$area_code)
```
```{r}
# Filter down to just Christchurch Central (area code 326600) and drop any rows where price is missing (can't include those in a median)
chch_central <- task1_data %>%
  filter(area_code == 326600, !is.na(price))

head(chch_central)
```
```{r}
# Check how many listings we actually got 
nrow(chch_central)
```
```{r}
# Calculate the median price
median_price <- median(chch_central$price)

median_price
```

```{r}
# Just for comparison - what's the median price across ALL of Christchurch, not just Central?
# if any rows have blank or missing values (NA) for the price, ignore them and calculate the median with what's left
median(task1_data$price, na.rm = TRUE)
```
Summary 21/9/2026 - Alice
Median airbnb price in CHCH central (area code 326600) is $239 (based on 724 listings with a recorded price)



# In which part of Christchurch can we observe the craziest (largest) gap between short- and long-term rental prices? - Alisha 
```{r}
# Check data columns
glimpse(joined_data)
```

Adding price_gap and filtering data
```{r}
joined_data <- joined_data %>%
  mutate(
    rental_price_per_night_median = `Median Rent` / 7, #putting median price per week into per day
    price_gap = price - rental_price_per_night_median # calculating the price gap for each row which is Airbnb nightly price minus median long-term rent per night
  )

# filter out rows without data (eg NA)
gap_data <- joined_data %>%
  filter(!is.na(price), !is.na(`Median Rent`))

```


## suberb analysis (SA2)
Most areas have between 1 and 2000 properties. This makes sense as sime areas have much more air bnb listings than others. This makes sense as areas like the city will be much more likely to have listings than smaller futher out areas like Aranui.

Decided to only include areas with over 15 properties listed
```{r}
# Summarise gap by area 
gap_by_area <- gap_data %>%
  group_by(area_code, area_name) %>%
  summarise(
    median_gap = median(price_gap, na.rm = TRUE), #calculating median price gap for areas excluding NA
    n = n(), # stores number of listings in each group
    .groups = "drop" #remove grouping after summarising
  ) %>%
  arrange(desc(median_gap)) %>% # sorts areas from biggest to smallest price gap 
  filter(n >= 15) # excludes areas with too few listings for the median to be reliable

head(gap_by_area, 15)
```
Even after filtering, some SA2 suburb areas (like Holmwood) are small, obscure statistical areas rather than widely-known suburbs. Tried grouping by ward instead because wards are bigger areas with higher listing counts and are more recognisable:

## Ward-level analysis
```{r}
# Summarise gap by ward
gap_by_ward <- gap_data %>%
  group_by(neighbourhood) %>%
  summarise(
    median_gap = median(price_gap, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(median_gap))

head(gap_by_ward, 15)
```
Looking at the wards they look like more well known offical areas of Christchurch

```{r}
# Bar chart of median gap by ward
gap_by_ward %>%
  ggplot(aes(x = reorder(neighbourhood, median_gap), y = median_gap, fill = median_gap)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "#a8dadc", high = "#1d3557") +
  labs(x = "Ward", y = "Median price gap ($ per night)", title = "Short vs long-term rental price gap by ward")
```
**Answer:** 
At the ward level (larger, more reliable sample sizes), Central Ward shows the largest gap between short- and long-term rental prices at $144.57/night (n=6426). 

(At the smaller SA2 suburb level, Holmwood shows an even larger gap ($217.21/night, n=74), but it's a small statistical area rather than a widely-recognised suburb.)

# Compare how many AirBnBs and rental properties we have in each location - Charlie
```{r}
# Compare Airbnb listing counts vs rental bond counts by area
comparison <- joined_data %>%
  group_by(area_code, area_name) %>%
  summarise(
    n_airbnb = n_distinct(id), # count of unique Airbnb listings in this area
    n_rental = mean(`Total Bonds`, na.rm = TRUE),  # already a per-area count, so average across quarters rather than sum
    .groups = "drop"
  ) %>%
  arrange(desc(n_airbnb)) # sort areas by number of Airbnb listings

comparison
```

```{r}
# Bar chart comparing Airbnb listings vs rental bonds for the top 10 areas by Airbnb count
comparison %>%
  filter(!is.na(n_rental)) %>% # drop areas with no rental bond data
  slice_max(n_airbnb, n = 10) %>% # keep only the top 10 areas by Airbnb listing count
  pivot_longer(cols = c(n_airbnb, n_rental), names_to = "type", values_to = "count") %>%
  ggplot(aes(x = reorder(area_name, count), y = count, fill = type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = NULL, y = "Count", fill = "Type",
       title = "Top 10 areas by Airbnb listings — AirBnB vs rental") +
  scale_fill_manual(values = c("n_airbnb" = "steelblue", "n_rental" = "orange"),
                     labels = c("n_airbnb" = "AirBnB listings", "n_rental" = "Rental bonds"))
```
```{r}
# Bar chart comparing Airbnb listings vs rental bonds for the top 10 areas ranked by Airbnb count
comparison %>%
  filter(!is.na(n_rental)) %>% # drop areas with no rental bond data
  slice_max(n_rental, n = 10) %>% # keep only the top 10 areas by Airbnb listing count
  pivot_longer(cols = c(n_airbnb, n_rental), names_to = "type", values_to = "count") %>%
  ggplot(aes(x = reorder(area_name, count), y = count, fill = type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = NULL, y = "Count", fill = "Type",
       title = "Top 10 areas by rental bonds — AirBnB vs rental") +
  scale_fill_manual(values = c("n_airbnb" = "steelblue", "n_rental" = "orange"),
                     labels = c("n_airbnb" = "AirBnB listings", "n_rental" = "Rental bonds"))
```

```{r}
# Bar chart comparing Airbnb listings vs rental bonds across ALL areas (not just top 10), sorted by combined total
comparison %>%
  filter(!is.na(n_rental)) %>%
  pivot_longer(cols = c(n_airbnb, n_rental), names_to = "type", values_to = "count") %>%
  ggplot(aes(x = reorder(area_name, count, sum), y = count, fill = type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = NULL, y = "Count", fill = "Type",
       title = "AirBnB listings vs rental bonds — all areas")
```

