---
title: "week_8_lab_airbnb"
author: "CHAKAR"
date: "2026-09-07"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Week 8 instruction for Airbnb file: All_chch listings (separate datasets
from oct 2025 to jun 2026 combined, but only for chch city). clean the
dataset -\> ultimate aim is going to be comparing rental prices and
number of available properties with tenancy dataset

```{r}
# load data as "data"
library(tidyverse)

data <- read_csv("week_8_datasets/all_chch_listings.csv")

# summary statistics for all columns
print(summary(data))
```

data has: Rows: 28795 Columns: 19

```{r}
# datatype
print(glimpse(data))
```

```{r}
# number of missing NA values
print(colSums(is.na(data)))
```

```{r}
# plot price distribution
ggplot(data, aes(x = price)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(title = "Distribution of Rental Prices", x = "Price", y = "Count")
```

Instruction from week 5 lab - so probably need days_since_review
column - added that column in

```{r}
# days since last reivew (calculate the time difference between scrape/publish date) then plot it
# added scrape_date, days_since_review
data <- data %>%
  mutate(
    scrape_date = ymd(paste0(month_year, "-01")), # Sets scrape date to the 1st of each month
    last_review = ymd(last_review),              # Converts the original text column directly into a Date
    days_since_review = as.numeric(scrape_date - last_review)
  )

head(data)

ggplot(data, aes(x = days_since_review)) +
  geom_histogram(bins = 50, fill = "forestgreen", color = "white") +
  theme_minimal() +
  labs(title = "Days Since Last Review", x = "Days", y = "Count")

# Find number of negative days
negative_rows <- data %>% 
  filter(days_since_review < 0)
negative_rows
```

Summary statistics: price: median price is \$205, but the max is
\$43,654 (a massive extreme outlier? error?). Price histogram: right
skewed, due to extreme price outliers

license: entirely NA (28,795 missing values) -\> consider dropping if
doesn't need this column

Missing values: price has 10,667 missing values: how to handle? or
filter them out?

last_review and reviews_per_month both have 2,627 missing values

days_since_review has 9418 negative values. Days since last review
histogram: Most reviews happened very recently (clumped near 0 days),
which makes sense for active listings, but there are long-tail stale
listings stretching out past 1,000+ days. For month_year column, change
values to include the day the scrape date was done (from the website).

```{r}
data_1 <- data %>%
  mutate( 
    # Assign the exact scrape date based on the month
    scrape_date = case_when(
      month_year == "2025-10" ~ ymd("2025-10-05"),
      month_year == "2025-11" ~ ymd("2025-11-07"),
      month_year == "2025-12" ~ ymd("2025-12-11"),
      month_year == "2026-01" ~ ymd("2026-01-16"),
      month_year == "2026-02" ~ ymd("2026-02-13"),
      month_year == "2026-03" ~ ymd("2026-03-17"),
      month_year == "2026-04" ~ ymd("2026-04-16"),
      month_year == "2026-05" ~ ymd("2026-05-23"),
      month_year == "2026-06" ~ ymd("2026-06-19"),
      TRUE ~ NA_Date_
    ),
    # Convert last_review into a proper Date type directly (no extra column needed!)
    last_review = ymd(last_review),
    # Calculate days since review using the updated last_review column
    days_since_review = as.numeric(scrape_date - last_review)
  )

# Check how many rows in days_since_review still have negative values
negative_rows <- data_1 %>%  
  filter(days_since_review < 0)
print(negative_rows)
```

Even after matching the scrape dates with the exact publication times,
1745 negative values remain in days_since_review column. (So these
reviews were posted in the days right after each monthly data
collection, counted as active listing. So leave these negative days for
now)

DEAL with price column first -\> extreme outliers + 10667 missing values
Check Price inconsistencies

```{r}
# Check the data type of the price column
class(data_1$price)

# 3. Check how many missing values (NAs) exist in the price column
sum(is.na(data_1$price))
```

See if 10667 missing prices are random or linked to specific months

```{r}
# Check missing prices grouped by each month scrape
missing_by_month <- data_1 %>%
  group_by(month_year) %>%
  summarise(
    total_listings = n(),
    missing_prices = sum(is.na(price)),
    pct_missing = (missing_prices / total_listings) * 100
  )

print(missing_by_month)
```

100% price missing values in 2025-12, 2026-01, 2026-02 months

checking whether a listing has only one unique price across all its
available months. If it does, it's stable, can use that price to fill in
December, January, and February.

```{r}
# Check if listing names change across months for the same ID
name_check <- data_1 %>%
  filter(!is.na(name)) %>%
  group_by(id) %>%
  summarise(unique_names = n_distinct(name))

# See how many listings ever changed their title
table(name_check$unique_names > 1)
```

FALSE (3,535 listings): For the vast majority of properties, the host
never changed the listing title across the entire time period you
observed. Their title stayed 100% identical.

TRUE (582 listings): For 582 properties, the host did change their
listing title at least once during your dataset

```{r}
# 1. Check price consistency for each listing across available months
price_consistency <- data_1 %>%
  filter(!is.na(price)) %>% # throw away rows where price is missing
  group_by(id) %>% # group all rows of individual listings based on id
  summarise( # each listing id becomes a single row
    unique_prices = n_distinct(price), # Counts how many different prices a listing had
    consistent_price = first(price),   # Grabs the first price it sees for that listing
    months_observed = n() # count how many total valid monthly data points exist for that specific listing in the dataset
  )
  
# Impute prices ONLY for listings that have a proven stable price history
# See how many listings have stable (unchanging) prices vs changing prices
table(price_consistency$unique_prices)

# Create a lookup table of ONLY the strictly stable listings
stable_price_lookup <- price_consistency %>%
  filter(unique_prices == 1) %>%
  select(id, consistent_price) 

# Impute prices safely using ifelse() to prevent type mismatch crashes
data_2 <- data_1 %>%
  left_join(stable_price_lookup, by = "id") %>%
  mutate(
    # If price is missing AND it matches the target months, pull from consistent_price
    price = ifelse(
      is.na(price) & month_year %in% c("2025-12", "2026-01", "2026-02"), 
      consistent_price, 
      price
    )
  ) %>%
  select(-consistent_price) # clean up temporary lookup column

# Check how many missing prices remain
print(sum(is.na(data_2$price)))
```

So 509 prices were filled as they were the same across every single
other months they appeared in (id)

Still has 10158 missing prices

Manage price outliers/extreme values Inspect the extreme rows

```{r}
# Check the top 20 most expensive listings using summary-file columns
top_expensive <- data_2 %>%
  filter(!is.na(price)) %>%
  arrange(desc(price)) %>%
  select(id, name, price, room_type, minimum_nights) %>%
  head(20)

print(top_expensive)
```

Rows 1-8 -\> likely error (checking latitude/altitude), doesn't make
sense to have that much a night Rows 10-15 -\> also likely error. priced
\>5000 per night in a private room doesn't make sense Rows 16-20..
maybe? entire home

Keep rows 1-15 will cause statistical bias into analysis. It distorts
the mean and standard deviation DECISION: Filter these 15 rows (note:
filter() keeps every row where the condition evaluates to TRUE, and
throws away anything that evaluates to FALSE)

```{r}
# Clean the dataset by removing anything $5,000 or higher (while keeping NAs. Otherwise R will remove NAs as well)
data_3 <- data_2 %>%
  filter(price < 5000 | is.na(price))

# Check how many rows were dropped
nrow(data_2) - nrow(data_3)
```

```{r}
# 1. Check the new maximum price to ensure the errors are truly gone
max(data_3$price, na.rm = TRUE)

# 2. Calculate clean summary statistics for prices
price_summary <- data_3 %>%
  filter(!is.na(price)) %>%
  summarise(
    mean_price = mean(price),
    median_price = median(price),
    min_price = min(price),
    max_price = max(price)
  )

print(price_summary)
```

What to do with remaining PRICE NAs? - leave them as NAs for now as not
sure what is happening downward

License 100% NAs - no universal registration rules so hosts didn't have
to register for official registration code to list their property -\>
DECISION DROP license column

```{r}
# Drop the useless license column
df_cleaned <- data_3 %>%
  select(-any_of("license"))

# check columns
colnames(df_cleaned)
```

check if reviews_per_month and last_review are related

```{r}
# Check if the exact same rows are missing values for both columns
identical(is.na(df_cleaned$reviews_per_month), is.na(df_cleaned$last_review))
```

if property has 0 reviews, both reviews_per_month, and last_review are
missing.

```{r}
# 1. Check if there are any literal 0 values already in reviews_per_month
sum(df_cleaned$reviews_per_month == 0, na.rm = TRUE)

# 2. Check if there are any non-NA values in last_review when reviews_per_month is NA (or vice versa)
df_cleaned %>%
  filter(is.na(reviews_per_month) != is.na(last_review)) %>%
  select(id, name, reviews_per_month, last_review)
```

There are zero listings.. every single listing either has a positive
review or is missing (NA). So reviews_per_month and last_review are 100%
structurally locked

DECISION: takes every missing value (NA) in the reviews_per_month column
and turns it into a literal 0

```{r}
# Finalize the review columns treatment based on structural verification
df_cleaned <- df_cleaned %>%
  mutate(
    # Recode NA to 0 for numerical analysis (since we proved 0 doesn't otherwise exist)
    reviews_per_month = replace_na(reviews_per_month, 0)
    # last_review remains NA, perfectly justified by the zero-review status
  )

# 1. Check how many NA values remain in reviews_per_month (should be 0)
sum(is.na(df_cleaned$reviews_per_month))

# 2. Check how many 0s are now in reviews_per_month (should match your previous missing count, 2627)
sum(df_cleaned$reviews_per_month == 0)
```

```{r}
# check columns
colSums(is.na(df_cleaned))
```

The Review-Related Columns (2,619 NAs) last_review: 2,619
last_review_date: 2,619 days_since_review: 2,619 These three columns are
locked together Decision to leave them as NA since can't calculate "days
since a review" if there has never been a review.

Price column: 10158 NAs

minimum_nights: 37 NAs

TO DECIDE: MINIMUM_NIGHTS NAs ?impute with median : min-1, max-365,
median-1, mean - 1.788

COLUMNS to drop: neighbourhood group only has 1 uniqe value so can't
help compare anything

```{r}
# Check is the right column
head(df_cleaned$neighbourhood_group)
```

DECISION: DROP neighbourhood_group column

```{r}
# Drop the zero-variance column (replace with the exact column name, e.g., neighbourhood_group)
df_cleaned <- df_cleaned %>%
  select(-neighbourhood_group)
colnames(df_cleaned)
```

```{r}
#check to see if last_review and last_review_date contain the exact same data
# Check if both columns are identical element-by-element
identical(df_cleaned$last_review, df_cleaned$last_review_date)
```

DECISIONS: Drop: host_name : irrelevant to analysis, and is personal
identifiable information

```{r}
df_cleaned <- df_cleaned %>%
  select(-c(host_name))
colnames(df_cleaned)
```

```{r}
#Check if any listing has more LTM reviews than total reviews (should output 0 rows)
df_cleaned %>%
  filter(number_of_reviews_ltm > number_of_reviews)
```

Check duplicates

```{r}
# Count how many fully duplicated rows exist
sum(duplicated(df_cleaned))

# Count how many duplicate IDs exist
sum(duplicated(df_cleaned$id))

#See which specific IDs are duplicated and how many times
df_cleaned %>%
  count(id) %>%
  filter(n > 1)

#check if these duplicate IDs are just the same property showing up in different months
# Pick one of the duplicated IDs from your output and see its monthly breakdown
df_cleaned %>% 
  filter(id == 204050) %>% # replace 204050 with one of your actual duplicated IDs
  select(id, month_year, price, availability_365)
```

The same listing IDs appear multiple times because Inside Airbnb scraped
them across different months, capturing how their prices and
availability dynamically changed over time.

This means they are not accidental duplicate errors. They are
time-series records.

```{r}
# 1. Quick check of column types (ensure prices/lat/long are numeric, dates are proper, etc.)
glimpse(df_cleaned)

# 2. Confirm total rows remaining
nrow(df_cleaned)
```

Summary:

1. price column 
Initially had 10,667 missing values, with 100%
missing values for the 3 months: Dec 2025–Feb 2026. Imputed 509 prices
where the listing had the same price across all other months. Removed 15
extreme outliers with prices > $5,000, which were likely errors.
Before: 28,795 rows After: 28,780 rows 
Question: What should we do with the remaining 10,158 missing values? The price column is still right-skewed.

2.  days_since_review column. 
Added days_since_review as instructed in Week 5. Initially, scrape_date was set to the first day of each month (YYYY-MM-01), resulting in 9,418 negative values. Updated scrape_date to use the actual day the listing was scraped and recalculated days_since_review. There are now 1,745 negative values, likely because some reviews were posted after the scrape date.
Question: Do we need this column? If so, how should we handle the
negative values?

3.  Removed columns:
license – removed because it was 100% missing.
neighbourhood_group – removed because all values were Christchurch_city, so it provides no useful comparison. 
host_name – removed because it contains personally identifiable information and is not relevant to the analysis.

4.  reviews_per_month and last_review:
Converted 2,619 missing values in reviews_per_month to 0, since listings with no reviews have both reviews_per_month and last_review recorded as NA. Left last_review NAs unchanged for now, since replacing them with 0 would not make sense for a date variable. 

5.  minimum_nights: 
Has 37 missing values. Current statistics: min = 1, median = 1, mean = 1.788, max = 365. Have left as it is. Depend on analysis later - impute missing values using median (1) or leave as it is?

6. Other findings:
There were no exact duplicates.
The same listing IDs appear multiple times because Inside Airbnb scraped
them across different months, capturing how their prices and
availability changed over time. So they are time-series records.

Before cleaning- number of rows: 28795, columns: 19
After cleaning: 28780, columns: 18
    

(Note: Cleaned dataset: df_cleaned)

```{r}
write.csv(df_cleaned, file = "week_9_datasets/airbnb_cleaned_week_8.csv", row.names = FALSE)
```
